---
title: "Introduction to lisaClust"
date: "`r BiocStyle::doc_date()`"
author:
- name: Ellis Patrick
  affiliation: School of Mathematics and Statistics, and Westmead Institute for Medical Research, University of Sydney, Australia
- name: Nicolas Canete
  affiliation: Westmead Institute for Medical Research, University of Sydney, Australia
- name: Alex Run Qin
  affiliation: School of Mathematics and Statistics, University of Sydney, Australia
package: "`r BiocStyle::pkg_ver('lisaClust')`"
abstract: >
  Which parts of a tumour are tumour cells, which are immune infiltrate and which are stroma? Which local
  neighbourhoods of cells are more common in one group of patients than another? lisaClust describes, for every
  cell, which cell types surround it, and groups cells with similar surroundings into regions. With a few regions
  these are tissue domains; with many they are cellular niches. This vignette finds domains in triple-negative
  breast cancers and niches in colorectal cancers, and compares them between groups of patients.
vignette: >
  %\VignetteIndexEntry{Introduction to lisaClust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc_float: true
    toc_depth: 2
bibliography: REFERENCES.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE, fig.retina = 1,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

```{css, echo = FALSE}
.callout { border-left: 4px solid; border-radius: 4px; padding: 0.6em 1em; margin: 1em 0; }
.callout > p:last-child { margin-bottom: 0; }
.callout-note { border-color: #1f6fb4; background: #eef4fb; }
.callout-tip { border-color: #2e7d32; background: #eef7ee; }
.callout-important { border-color: #b3261e; background: #fbeeee; }
```

# Introduction

For each cell, lisaClust [@patrick2023] counts the cells of every type within a few radii of it and compares each count with
what would be expected if that type were spread evenly over the image. These local indicators of spatial
association (LISA) describe the cell's surroundings: a tumour cell deep in a tumour nest has many tumour neighbours
and few immune neighbours, while a tumour cell at the invasive margin has both. Clustering the cells by their
LISA groups cells with similar surroundings, whatever their own type, into **regions**.

The number of regions sets the scale of the question. With a few regions, they are broad tissue **domains**, such
as tumour, stroma and immune infiltrate. With 20 or more, they are cellular **niches**: recurring local
combinations of cell types, such as T cells mixed with macrophages at a tumour border.

lisaClust needs only the type and position of every cell, so it applies to any technology that gives segmented,
typed cells: spatial proteomics such as imaging mass cytometry, MIBI and CODEX, and spatial transcriptomics such
as Xenium, CosMx, MERSCOPE or high-definition spatial transcriptomics [@patrick2023]. It is not designed for
spot-based data such as Visium, where a spot holds several cells. It works alongside other SydneyBioX packages: *spicyR* tests whether pairs
of cell types co-localise differently between groups of patients, and *Statial* tests co-localisation within
tissue compartments. Here we use spicyR's `colTest()` to compare regions between groups of patients, and
*tidySpatialExperiment* to summarise the cells of each image with dplyr verbs.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("lisaClust", "spicyR", "tidySpatialExperiment", "SpatialDatasets"))
```

```{r packages}
library(lisaClust)
library(spicyR)
library(tidySpatialExperiment)
library(SpatialDatasets)
library(ggplot2)
theme_set(theme_classic())
```

# Tissue domains in triple-negative breast cancer

## The data

@keren2018 imaged tumours from 40 patients with triple-negative breast cancer by MIBI-TOF, one image per patient,
and classified each tumour by how its immune and tumour cells are arranged: *compartmentalised*, where immune
cells and tumour cells occupy separate areas; *mixed*, where they are interspersed; and *cold*, with few immune
cells. The data are a `SpatialExperiment` in the SpatialDatasets package, with the coordinates in pixels.

```{r keren-data}
kerenSPE <- spe_Keren_2018()
kerenSPE |> distinct(imageID, tumour_type) |> count(tumour_type)
sort(table(kerenSPE$cellType), decreasing = TRUE)
```

## Finding domains

`lisaClust()` computes the LISA of every cell and clusters them with k-means into `k` regions, stored in a new
`colData` column (`region` by default). It looks for columns called `imageID` and `cellType`, and takes the
coordinates of a `SpatialExperiment` from `spatialCoords()`. Here we ask for four domains, describing each cell's
surroundings within 20, 50 and 100 pixels.

```{r keren-lisaclust}
set.seed(51773)
kerenSPE <- lisaClust(kerenSPE, k = 4, r = c(20, 50, 100))
table(kerenSPE$region)
```

k-means starts from random centres, so set a seed to make the regions reproducible.

## What is in each domain?

`regionMap()` shows how much more often each cell type is found in each region than if cell types were spread
evenly over the regions: above one, the type is enriched in that region.

```{r keren-region-map, fig.width = 5, fig.height = 5}
regionMap(kerenSPE, type = "bubble")
```

The regions are numbered arbitrarily. `nameRegions()` names each one by the marker cell type it is most
enriched for, here among tumour cells, CD8 T cells, B cells and mesenchymal cells. Regions most enriched for the
same marker get the same name, so they are merged into one domain.

```{r keren-domain-names}
markers <- c(tumour = "Keratin_Tumour", `T cell` = "CD8_T_cell", `B cell` = "B_cell", stroma = "Mesenchymal")
kerenSPE <- nameRegions(kerenSPE, markers, regionName = "domain")
round(100 * prop.table(table(kerenSPE$domain)), 1)
```

The cell types in each domain:

```{r keren-domain-map, fig.width = 4.5, fig.height = 5}
regionMap(kerenSPE, region = "domain", type = "bubble")
```

## Domains in two tumours

`hatchingPlot()` draws the cells coloured by type, with each region outlined and filled with its own hatching, so
types and regions can be seen together. Here are the compartmentalised tumour and the mixed tumour with the most
cells in the T cell domain. With tidySpatialExperiment, `group_by()` and `summarise()` work on the cells of a
`SpatialExperiment` and return a table.

```{r keren-examples}
tcellShare <- kerenSPE |>
  group_by(imageID, tumour_type) |>
  summarise(share = mean(domain == "T cell"), .groups = "drop")
examples <- tcellShare |>
  filter(tumour_type != "cold") |>
  group_by(tumour_type) |>
  slice_max(share)
examples
```

```{r keren-hatching, fig.width = 10, fig.height = 5.5}
hatchingPlot(kerenSPE, useImages = examples$imageID, region = "domain")
```

## Do the domains differ between kinds of tumour?

Each patient has one image, so the share of each domain in an image is a value per patient. `regionBoxPlot()`
draws these shares by group.

```{r keren-domain-share, fig.width = 8, fig.height = 4}
regionBoxPlot(kerenSPE, condition = "tumour_type", region = "domain")
```

spicyR's `colTest()` computes the share of each domain in each image and compares the groups with a Wilcoxon
test; here we compare compartmentalised with mixed tumours.

```{r keren-domain-test}
domainTests <- kerenSPE |>
  filter(tumour_type != "cold") |>
  colTest(condition = "tumour_type", feature = "domain", type = "wilcox")
domainTests
medians <- with(tcellShare, tapply(share, tumour_type, median))
round(100 * medians, 1)
pT <- domainTests["T cell", "pval"]
```

The T cell domain makes up a median of `r round(100 * medians["compartmentalised"], 1)`% of the cells in a
compartmentalised tumour and `r round(100 * medians["mixed"], 1)`% in a mixed one (Wilcoxon p =
`r format(pT, digits = 2)`). `r if (medians["compartmentalised"] > medians["mixed"]) "In compartmentalised tumours the immune cells form their own areas, which become a domain; in mixed tumours they sit among the tumour cells and fall into the tumour domain." else ""`

::: {.callout .callout-important}
@keren2018 classified the tumours from how their immune and tumour cells are arranged, so this difference is a
check that the domains capture that arrangement, not an independent finding.
:::

# Cellular niches in colorectal cancer

## The data

@schurch2020 imaged colorectal cancers from 35 patients by CODEX, four tissue microarray cores per patient. The
patients had one of two patterns of immune response: a Crohn's-like reaction (CLR, with lymphoid follicles; 17
patients) or diffuse inflammatory infiltration (DII; 18 patients). We leave out the objects the authors labelled
as debris (`dirt`).

```{r schurch-data}
schurchSPE <- spe_Schurch_2020() |>
  filter(cellType != "dirt") |>
  mutate(condition = ifelse(group == 1, "CLR", "DII"))
schurchSPE |> distinct(patients, condition) |> count(condition)
length(unique(schurchSPE$cellType))
```

## Finding niches

With 20 regions, the regions are finer combinations of cell types. We store them in a column called `niche`.

```{r schurch-lisaclust}
set.seed(51773)
schurchSPE <- lisaClust(schurchSPE, k = 20, r = c(20, 50, 100), regionName = "niche")
```

```{r schurch-region-map, fig.width = 9, fig.height = 7}
regionMap(schurchSPE, region = "niche", type = "bubble") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Comparing niches between groups of patients

The question is whether a niche makes up more of the tissue in one group of patients than in the other. The four
cores of a patient come from the same tumour, so they are not independent: we pool the cells of each patient's
cores, so that each patient contributes one share of each niche, and compare the two groups of patients. Setting
`imageID = "patients"` in `colTest()` does this pooling. With 20 niches tested, `colTest()` also adjusts the
p-values for multiple testing (`adjPval`).

```{r schurch-patient-test}
patientTests <- colTest(schurchSPE, condition = "condition", feature = "niche",
                        imageID = "patients", type = "wilcox")
head(patientTests, 5)
```

For comparison, the same test treating each core as an independent sample:

```{r schurch-core-test}
coreTests <- colTest(schurchSPE, condition = "condition", feature = "niche",
                     imageID = "imageID", type = "wilcox")
c(patients = sum(patientTests$adjPval < 0.05), cores = sum(coreTests$adjPval < 0.05))
```

```{r schurch-counts, include = FALSE}
nPatients <- sum(patientTests$adjPval < 0.05)
nCores <- sum(coreTests$adjPval < 0.05)
niches <- function(n) if (n == 1) "1 niche differs" else paste(n, "niches differ")
```

With the 35 patients as the units, `r niches(nPatients)` between CLR and DII patients at a 5% false discovery rate
(smallest adjusted p-value `r format(min(patientTests$adjPval), digits = 2)`). Treating each of the 140 cores as an
independent sample instead, `r niches(nCores)`, with a smallest adjusted p-value of
`r format(min(coreTests$adjPval), digits = 2)`.`r if (nCores > nPatients) " The extra niches are not supported by the patients." else ""`

::: {.callout .callout-important}
Cores or images from the same patient are alike, so treating them as independent overstates the evidence
and can turn up differences that the patients do not support. Compare niches with patients as the units, and adjust
for the number of niches tested.
:::

`regionMap(type = "table")` gives the enrichment of each cell type in each niche as a matrix, so we can list
the cell types the niche with the smallest p-value is most enriched for, and its median share in each group.

```{r schurch-top-niche-types}
top <- patientTests$cluster[1]
enrichment <- regionMap(schurchSPE, region = "niche", type = "table")
topTypes <- sort(enrichment[, top], decreasing = TRUE)
round(topTypes[1:5], 1)

topMedians <- schurchSPE |>
  group_by(patients, condition) |>
  summarise(share = mean(niche == top), .groups = "drop") |>
  with(tapply(share, condition, median))
round(100 * topMedians, 1)
```

The niche with the smallest p-value, `r top`, is most enriched for
`r paste(names(topTypes)[1:3], collapse = ", ")`. It makes up a median of
`r round(100 * topMedians["CLR"], 1)`% of the cells in CLR patients and
`r round(100 * topMedians["DII"], 1)`% in DII patients`r if (names(topTypes)[1] == "b_cells" && topMedians["CLR"] > topMedians["DII"]) ", as expected for the lymphoid follicles that define a Crohn's-like reaction" else ""`:

```{r schurch-top-niche, fig.width = 3, fig.height = 4}
regionBoxPlot(schurchSPE, condition = "condition", region = "niche", imageID = "patients", regions = top)
```

# Using your own clustering

`lisa()` returns the LISA themselves, one row per cell and one column per radius and cell type, so you can
cluster them any way you like.

```{r lisa-curves}
curves <- lisa(kerenSPE[, kerenSPE$imageID %in% examples$imageID], r = c(20, 50, 100))
dim(curves)
curves[1:3, 1:4]

# for example, k-means with several random starts
set.seed(51773)
km <- kmeans(curves, centers = 4, nstart = 10, iter.max = 50)
table(km$cluster)
```

# Choosing the settings

::: {.callout .callout-tip}
**The number of regions.** A few regions describe the large compartments of a tissue; many regions separate finer
neighbourhoods but split rarer combinations across several regions and make each one harder to interpret. Look at
`regionMap()` for a few values of `k` and choose the coarsest one that separates the structures you care about;
generic rules for the number of clusters, such as the elbow or silhouette, can merge tissue structures that are
known to be distinct [@patrick2023].
:::

::: {.callout .callout-tip}
**The radii.** Choose them from the scales of interaction you expect, in the units of the coordinates: a radius of
two or three cell diameters describes the cell's immediate neighbours, and larger radii describe its wider
surroundings. Several radii (`r`) are used together; run
time grows with the number and size of the radii, so a few well-chosen ones are better than many.
:::

::: {.callout .callout-note}
**Uneven cell density.** By default the expected count assumes cells are spread evenly over the image. With
`sigma`, each neighbour is weighted by the inverse of the local density of all cells, estimated with a Gaussian
kernel of bandwidth `sigma`, so that a dense area does not by itself look enriched for every cell type.
:::

# How it works

For cell *i*, cell type *j* and radius *r*, let *n<sub>ij</sub>(r)* be the number of cells of type *j* within *r*
of cell *i*, counting cell *i* itself when it is of type *j*. If the cells of type *j* were spread evenly over the image window *W*, with density
$\lambda_j = N_j / |W|$, the expected count would be

$$E_{ij}(r) = \lambda_j \, \pi r^2 e_i(r),$$

where $e_i(r)$ is the share of the disc of radius *r* around cell *i* that lies inside the window, which corrects
for cells near the edge. The LISA is the standardised difference

$$\frac{n_{ij}(r) - E_{ij}(r)}{\sqrt{E_{ij}(r)}},$$

a local version of Ripley's K function for the pair (`lisaFunc = "L"` gives
$\sqrt{n_{ij}(r)} - \sqrt{E_{ij}(r)}$ instead). The window is the convex hull of the image's cells by default
(`window`). The neighbour search and the edge correction run in C++, so the LISA of hundreds of thousands of cells
take seconds.

# Reporting results

A methods sentence might read: "We used lisaClust (version `r packageVersion("lisaClust")`) to compute, for each
cell, local indicators of spatial association with every cell type at radii of 20, 50 and 100 pixels, and
clustered the cells into 20 niches by k-means. The share of each niche in each patient was compared between
groups with a Wilcoxon test (spicyR's colTest), with patients as the units, and p-values were adjusted across niches by the
Benjamini–Hochberg method." Show `regionMap()` and a `hatchingPlot()` of example images with the results.

```{r citation}
citation("lisaClust")
```

# Session information

```{r session}
sessionInfo()
```

# References
